\documentclass[12pt]{article}

\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{titling}
\setlength{\droptitle}{-3em}

\lstset{
    language=Python,
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue},
    commentstyle=\color{green!60!black},
    stringstyle=\color{red},
    numbers=left,
    numberstyle=\tiny\color{gray},
    frame=single,
    backgroundcolor=\color{gray!10},
    breaklines=true,
    showstringspaces=false
}

\geometry{margin=1in}

\title{ODE Project Phase 1}
\author{
Greyson Knippel \\[4pt]
MATH 260-1 --- Gonzaga University \\[4pt]
Professor Jensen
}
\date{\today}



\begin{document}

\maketitle

\section*{problem 2 (balloon)}


A balloon is ascending at a rate of 15 m/s at a height of 100 m above the ground when a package
is dropped from the gondola. How long will it take the package to reach the ground? Ignore air resistance.

\subsection*{known variables}
when the package is dropped, it has an upward velocity of the balloon.

\begin{center}
\begin{tabular}{ll}
    \hline
    variable & value \\
    \hline
    initial height $y_0$ & $100$ m \\
    initial velocity $v_0$ & $+15$ m/s (upward) \\
    acceleration $a$ & $-9.8$ m/s$^2$ (gravity) \\
    final height $y$ & $0$ m (ground) \\
    \hline
\end{tabular}
\end{center}

\subsection*{setting up the equation}
using the kinematic equation for position:
\[
y = y_0 + v_0 t + \frac{1}{2}at^2
\]
substituting known values and setting $y = 0$:
\[
0 = 100 + 15t + \frac{1}{2}(-9.8)t^2
\]
\[
0 = 100 + 15t - 4.9t^2
\]
rearranging into quadratic form $at^2 + bt + c = 0$:
\[
4.9t^2 - 15t - 100 = 0
\]
where $a = 4.9$, $b = -15$, $c = -100$

\subsection*{solving with quadratic formula}
\[
t = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
\]
calculate the discriminant:
\[
b^2 - 4ac = (-15)^2 - 4(4.9)(-100) = 225 + 1960 = 2185
\]
substituting:
\[
t = \frac{15 \pm \sqrt{2185}}{9.8} = \frac{15 \pm 46.74}{9.8}
\]
gives us two solutions:
\[
t = \frac{15 + 46.74}{9.8} \approx 6.3 \text{ s} \qquad \text{or} \qquad t = \frac{15 - 46.74}{9.8} \approx -3.24 \text{ s}
\]

\subsection*{result}
since time can't be negative, we don't care about $t \approx -3.24$ s. the package reaches the ground in:
\[
\boxed{t \approx 6.3 \text{ seconds}}
\]

\section*{problem 4 (rocket)}

\subsection*{}
A rocket is fired vertically and ascends with constant acceleration $a = 100$ m/s$^2$ for 1.0 min.
At that point, the rocket motor shuts off and the rocket continues upward under the influence of gravity.
Find the maximum altitude acquired by the rocket and the total time elapsed from take-off until the
rocket returns to the earth. Ignore air resistance.

\subsection*{overview}
this problem has two parts:
\begin{itemize}
    \item \textbf{part 1:} engine ON — constant acceleration upward for 60 seconds
    \item \textbf{part 2:} engine OFF — only gravity acts on the rocket
\end{itemize}
the kinematic equations used are:
\begin{align}
    v &= v_0 + at \tag{1} \\
    y &= y_0 + v_0 t + \frac{1}{2}at^2 \tag{2} \\
    v^2 &= v_0^2 + 2a\Delta y \tag{3}
\end{align}

\subsection*{part 1: engine on ($t = 0$ to $t = 60$ s)}

known quantities:
\begin{center}
\begin{tabular}{ll}
    \hline
    variable & value \\
    \hline
    initial height $y_0$ & $0$ m \\
    initial velocity $v_0$ & $0$ m/s \\
    acceleration $a$ & $+100$ m/s$^2$ \\
    time $t$ & $60$ s \\
    \hline
\end{tabular}
\end{center}

\textbf{height at engine shutoff} (using equation 2):
\[
y_1 = 0 + (0)(60) + \frac{1}{2}(100)(60)^2 = 180{,}000 \text{ m}
\]

\textbf{velocity at engine shutoff} (using equation 1):
\[
v_1 = 0 + (100)(60) = 6{,}000 \text{ m/s}
\]

\subsection*{part 2: engine off}

the rocket enters part 2 with $y_0 = 180{,}000$ m, $v_0 = 6{,}000$ m/s, and $a = -9.8$ m/s$^2$.
this part is split into two (other) parts.

\subsubsection*{part 2a: coasting up to max altitude}

at maximum altitude, the rocket momentarily stops, so $v = 0$. using equation (3) to find the
additional height gained:
\[
0 = (6{,}000)^2 + 2(-9.8)\Delta y
\]
\[
\Delta y = \frac{(6{,}000)^2}{2(9.8)} = 1{,}836{,}734.94 \text{ m}
\]

\textbf{maximum altitude:}
\[
y_{\max} = 180{,}000 + 1{,}836{,}734.94 = \boxed{2{,}016{,}734.94 \text{ m}}
\]

time to reach maximum altitude (using equation 1):
\[
0 = 6{,}000 + (-9.8)t \implies t_{2a} = \frac{6{,}000}{9.8} \approx 612.24 \text{ s}
\]

\subsubsection*{part 2b: falling from max altitude to ground}
the rocket falls from rest at $y_0 = 2{,}016{,}734.94$ m. using equation (2) with $v_0 = 0$:
\[
0 = 2{,}016{,}734.94 - \frac{1}{2}(9.8)t^2
\]
\[
t^2 = \frac{2{,}016{,}734.94}{4.9} = 411{,}578.56
\]
\[
t_{2b} = \sqrt{411{,}578.56} \approx 641.23 \text{ s}
\]

\subsection*{total time}

\[
t_{\text{total}} = t_1 + t_{2a} + t_{2b} = 60 + 612.24 + 641.23 = \boxed{1{,}313.47 \text{ s}}
\]

\section*{guide for solving:}

\subsection*{background info}

\begin{quote}
what helped me solve these problems was the background information I gained from taking AP Physics in
high school. it took me a little while to remember some of the steps to use these formulas, but it
came back with that previous knowledge of kinematics.

\medskip

if you did \textbf{not} take AP Physics in high school, or just can't remember how to use these
equations, a simple google search or kinematics textbook can tell you how to use the equations
used in the problems to find velocity, time, acceleration, etc.
\end{quote}


\subsection*{mock report:}


\begin{quote}
both of these problems can be solved with fairly basic kinematics equations, splitting these
questions into different parts.


\end{quote}

\end{document}